---
title: "analysis"
author: "Jordan Hutchings"
date: "25/10/2021"
output: html_document
always_allow_html: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

The main question of interest is simply, how much money would you make from 
betting $100 on a given tour golfer. This analysis was motivated by the question, 
if you were to bet $100 on Tiger Woods in each major he competed in, how much 
money would you have today. 

## Game plan
1. Scrape betting lines for the Major Champions from here: 
2. Calculate payoffs per player
3. Visualizations

The majority of the work here will be to properly scrape the betting odds data 
and field lists for each of the majors in our dataset. 

Once we have the data, it will simply be a count of each player name, and the 
sum of their winning odds times $100. 

## Scraping the data 
We can use `rvest` to collect the names of the players on a webpage. Notice the 
webpages we are interested in share very similar URL structure. This allows us to 
rip through each year collecting the table elements and saving into data frames for 
easy analysis. 

```{r, warning=FALSE, message=FALSE, cache=TRUE}
pacman::p_load(rvest, dplyr, ggplot2, ggrepel)

# get player list for a given webpage ----
url <- "https://www.sportsoddshistory.com/golf-main/?y=2021&sa=golf&a=mast&o=r"
html <- read_html(url)
tbl <- html %>%
  html_table()
tbl[[2]]$Player

# collect all players in the masters ----
pre_url <- "https://www.sportsoddshistory.com/golf-main/?y="
post_url <- "&sa=golf&a=mast&o=r"

years <- seq(1985, 2021, 1) #1985 start 

get_data <- function(x){
  # download html webpage, take names from second table on the page, 
  # this will break if the table order differs - but it seems to be 
  # consistent based on checking some webpages. 
  cat("Collecting Masters ", x, "\n")
  Sys.sleep(runif(n = 1, 0.5, 1)) # hopefully this is pretty chill of a wait time
  path <- paste0(pre_url, x, post_url)
  html <- read_html(path)
  tbl <- html %>%
    html_table()
  player_list <- tbl[[2]]$Player
  
  player_list
}

masters_players <- lapply(years, get_data)
tbl <- unlist(masters_players, use.names=FALSE)
tbl <- data.frame(tbl)
names(tbl) <- "Player"

# collect masters champions and their historical odds ----

major_odds <- "https://www.sportsoddshistory.com/golf-champs/"
mo <- read_html(major_odds)
mo <- mo %>%
  html_table()

masters_odds <- data.frame(mo[1])
masters_odds <- masters_odds %>%
  select(Player = Player, Odds = Prior.to...) %>%
  filter(Player != "Player")


# merge datasets ----
masters_appearences <- tbl %>%
  group_by(Player) %>%
  summarise(n = n())

masters_results <- masters_odds %>%
  mutate(payoff = as.numeric(Odds) + 100) %>%
  group_by(Player) %>%
  summarise(payouts = sum(payoff)) %>%
  left_join(masters_appearences, by = "Player") %>%
  mutate(profit = payouts - 100 * n,
         ev = payouts / (100 * n))

```

## Masters Analysis 
I first used the Masters as a proof of concept. Below are some visualizations of 
the top players in terms of total profit. 


```{r, warning=FALSE, message=FALSE}
ggplot(masters_results, aes(x = n, y = profit, label = Player)) +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_point() + 
  geom_text_repel(hjust = -0.1, min.segment.length = 0.2, color = "darkgrey") + 
  theme_classic() + 
  labs(title = "Historical Payouts if you bet $100 on each player \nin the Masters (1985 - 2021)", 
       x = "Number of starts", 
       y = "Profit ($)")

masters_results$Player <- reorder(masters_results$Player, masters_results$profit)
masters_results %>%
  ggplot(aes(x=Player, y=profit)) + 
  geom_segment(aes(x=Player, xend=Player, y=0, yend=profit), size = 1, color = "#ffdf00", alpha = 0.3) + 
  geom_point(size = 3, color = "#076652") + 
  coord_flip() + 
  theme_classic(12) + 
  labs(title = "Total profit Earned by Placing a $100 bet on each \nPlayer in the maters (1985 - 2021)", 
        y = "Profit ($)", 
        x = "") + 
    geom_abline(xintercept = 0, color = "darkgrey", linetype = "dashed")



```
